---
title: "Graphing Random Forest and LANDIS Carbon"
author: "Wesley Rancher"
date: "2025-04-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("F:/jetstream/")

library(dplyr)
library(ggplot2)
```

Random forest by itself

```{r}
predictions <- read.csv("data/PredictedCarbon_RasterValues_04252025.csv")

predictions$Model <- "Random forest"

colors <- c("Alaskan birch" = "#67161CFF","Black spruce" = "#3F6148FF",
            "White spruce" = "#A4804CFF","Trembling aspen" = "#4B5F80FF")


predictions<- predictions%>%
  mutate(Year=as.numeric(Year))%>%
  mutate(Species = ifelse(Species == "Resin birch", "Alaskan birch", Species),
         Species = ifelse(Species == "Quaking aspen", "Trembling aspen", Species))

#needs grouping
predictions%>%
  #plotting
  ggplot(aes(x=Year, y=.pred, color = Species))+
  #geom_ribbon(aes(ymin = SumCarbon_gm2 - SdCarbon_gm2 , ymax = SumCarbon_gm2 + SdCarbon_gm2 , fill = Species), alpha = 0.4, color = NA)+
  geom_point()+
  geom_smooth(method = "loess", alpha=0.4)+
  #geom_line(linewidth=1)+
  scale_color_manual(values=colors)+
  scale_fill_manual(values = colors)+
  #scale_x_continuous(limits=c(2000,2025), breaks = 5)+
  facet_grid(Species~.)+
  theme_bw()+
  theme(legend.position = "none")
```


Add in landis

```{r}
#reference files to get regular points
ref <- rast("F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/Cleaned_Inputs/AK_ClimateMap_10_Regions.tif")
bounds <- vect("F:/randomForest_Landis/input/shp/FullLandscapeV3_082722.shp")

crs(ref) <- "EPSG:3338"
ref_cleaned <- mask(ref, bounds)
regularPoints <- spatSample(ref_cleaned, size=2000000, method = "regular", as.points = TRUE, ext = bounds, na.rm=TRUE)

#out patterns (the files I want)
years <- 0:24
species <- c("BlackSpruce", "WhiteSpruce", "QuakingAspen", "PaperBirch")
species_pattern <- paste(species, collapse = "|")
year_pattern <- paste0("-",years,".img", collapse = "|")
combined_pattern <- paste0("(", species_pattern, ").*(", year_pattern, ")")

#grab files
outdir <- "F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/TestRuns/interior_hist_ncar_04252025_GrowthOnly_AllSpp/biomass/"
biomass_files_one_rep <- list.files(outdir, pattern = combined_pattern, full.names=T)

#get average
get_biomass_average <- function(file){
  bname <- basename(file)
  species <- strsplit(bname, "-")[[1]][1]
  yr <- sub(".*-(\\d+)\\.img", "\\1", bname)
  
  #read the file in as a raster and extract values
  r <- rast(file)
  crs(r) <- "EPSG:3338"
  ext(r) <- ext(ref_cleaned)
  r <- flip(r, direction="vertical")
  r_clipped <- mask(r, bounds)
  r_w_ecoreg <- c(r_clipped,ref_cleaned)
  
  #summary stats
  r_vals <- terra::extract(r_clipped, regularPoints, xy = TRUE)%>%
      mutate(.pred=Layer_1*0.47)%>%select(-Layer_1)%>%
      slice(seq(1, 2000000, by = 2000))
      #group_by(AK_ClimateMap_10_Regions)%>%
      #rename(Ecoregion=AK_ClimateMap_10_Regions)%>%
      #summarise(AvgCarbon_gm2=mean(Layer_1*0.47, na.rm = TRUE),
      #          SdCarbon_gm2=sd(Layer_1*0.47, na.rm = TRUE))
      
  #add IDs
  r_vals$Species <- species
  r_vals$Year <- as.numeric(yr)+2000
  return(r_vals)
}

processed_biomass_maps <- lapply(biomass_files_one_rep, get_biomass_average)
landis_predictions <- do.call(rbind, processed_biomass_maps) %>% select(-ID)
landis_predictions$Model <- "LANDIS-II"

```

Plot together

```{r}

predictions$Model <- "Random forest"
master_df <- rbind(predictions, landis_predictions)%>%#use this df for any raster analysis to compare?
  mutate(Species = ifelse(Species == "PaperBirch", "Alaskan birch", Species),
         Species = ifelse(Species == "QuakingAspen", "Trembling aspen", Species),
         Species = ifelse(Species == "BlackSpruce", "Black spruce", Species),
         Species = ifelse(Species == "WhiteSpruce", "White spruce", Species))
summarised_df <- master_df %>%
  mutate(Year=as.numeric(Year)) %>%
  
  group_by(Year, Species, Model) %>%
  summarise(AvgCarbon_gm2=mean(.pred, na.rm=TRUE),
            SDCarbon_gm2=sd(.pred, na.rm=TRUE))

model_colors <- c("darkorange", "darkorchid")
#double line plot faceted by species
summarised_df %>%
  ggplot(aes(x=Year, y=AvgCarbon_gm2, color = Species, group = Model, linetype = Model))+
  #geom_ribbon(aes(ymin = AvgCarbon_gm2 - SDCarbon_gm2 , ymax = AvgCarbon_gm2 + SDCarbon_gm2 , fill = Model, group = Model), position = "identity", alpha = 0.4)+
  #geom_smooth()+
  geom_line(linewidth=1.25)+
  #geom_point()+
  scale_color_manual(values = colors)+
  scale_fill_manual(values = model_colors)+
  coord_cartesian(ylim = c(0, NA))+
  facet_grid(Species~., scales = "free")+
  theme_bw()+
  labs(y="Aboveground C (g m²)")+
  theme(legend.position = "none")+
  guides(color = "none", fill = "none")

#difference
library(tidyr)
diff_df <- summarised_df %>%
  select(Year, Species, Model, AvgCarbon_gm2) %>%
  pivot_wider(names_from = Model, values_from = AvgCarbon_gm2) %>%
  mutate(Diff = `LANDIS-II` - `Random forest`)

ggplot(diff_df, aes(x = Year, y = Diff, color = Species)) +
  geom_line(linewidth=1.2) +
  #geom_point()+
  scale_color_manual(values=colors)+
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(Species~., scales = "free_y") +
  labs(y = "Differenced Aboveground C (g m²)", x = "Year") +
  theme_bw()+
  theme(legend.position = "none")
```

Random forest comparison. Scripts for random forest on remote sensing and landis data are in this directory. From those scripts I output the variable importance scores for each species as dataframes that way I can make some faceted plots here.

1) variable importances
2) trends in those variables (ordination plot?)
3) model performance?? (just remote sensing?)

```{r}
library(patchwork)
#clean var imp output a little
varImp_rf <- read.csv("data/VariableImportance_04162025.csv")  %>%
  mutate(Model = "Random-forest")%>%
  mutate(Spp = ifelse(Spp == "resin birch", "Alaskan birch", Spp),
         Spp = ifelse(Spp == "quaking aspen", "Trembling aspen", Spp),
         Spp = ifelse(Spp == "black spruce", "Black spruce", Spp),
         Spp = ifelse(Spp == "white spruce", "White spruce", Spp))

varImp_landis <- read.csv("data/VariableImportance_LANDIS_04282025.csv") %>%
  mutate(Model = "LANDIS-II")%>%
  rename(Spp=Species)%>%
  mutate(Spp = ifelse(Spp == "PaperBirch", "Alaskan birch", Spp),
         Spp = ifelse(Spp == "QuakingAspen", "Trembling aspen", Spp),
         Spp = ifelse(Spp == "BlackSpruce", "Black spruce", Spp),
         Spp = ifelse(Spp == "WhiteSpruce", "White spruce", Spp))
  
#join
#varImp_df_main <- bind_rows(varImp_rf, varImp_landis) %>% mutate(Model=as.factor(Model))

#random forest
vip_rf <- varImp_rf%>%
  
  group_by(Spp) %>%
  slice_max(order_by = Importance, n = 7, with_ties = FALSE) %>%
  arrange(Spp, desc(Importance))%>%
  ungroup() %>%
  ggplot(aes(x=Importance, y=Variable, fill=Spp))+
  #geom_point()+
  geom_bar(stat="identity")+
  #geom_line(linewidth=1)+
  scale_color_manual(values=colors)+
  scale_fill_manual(values=colors)+
  #scale_x_continuous(limits=c(2000,2025), breaks = 5)+
  facet_grid(Spp~., scales = "free_y")+
  theme_bw()+
  #theme(axis.line=element_line(color = "black"))+
  labs(title = "Random forest", x="Feature importance")+
  theme(axis.text = element_text(color = "black", size = 10),
        axis.text.x = element_blank(),
        axis.title = element_blank(),
        axis.title.x = element_text(color = "black"),
        strip.text = element_text(color = "black", size = 11),
        legend.position = "none",
        panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_blank())

#instead of faceting I just make two plots and add
vip_land <- varImp_landis%>%
  
  group_by(Spp) %>%
  slice_max(order_by = Importance, n = 7, with_ties = FALSE) %>%
  arrange(Spp, desc(Importance))%>%
  ungroup() %>%
  ggplot(aes(x=Importance, y=Variable, fill=Spp))+
  #geom_point()+
  geom_bar(stat="identity")+
  #geom_line(linewidth=1)+
  scale_color_manual(values=colors)+
  scale_fill_manual(values=colors)+
  #scale_x_continuous(limits=c(2000,2025), breaks = 5)+
  facet_grid(Spp~., scales = "free_y")+
  theme_bw()+
  #theme(axis.line=element_line(color = "black"))+
  labs(title = "LANDIS-II", x="Feature importance")+
  theme(axis.text = element_text(color = "black", size = 10),
        axis.text.x = element_blank(),
        axis.title = element_blank(),
        axis.title.x = element_text(color = "black"),
        strip.text = element_text(color = "black", size = 11),
        legend.position = "none",
        panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_blank())

vip_rf | vip_land
```

proximity matrix results from the random forest

```{r}


ggplot(mds_df, aes(x = Dim1, y = Dim2, color = Species)) +
  geom_point() +
  theme_minimal()

```
