---
title: "Random Forest Future Climate"
author: "Wesley Rancher"
date: "2025-04-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
setwd("F:/jetstream/")
dir <- "F:/remoteSensing/data/output/"
```

Read in training dataset and rasters to make predictions on

Landsat data MSS, TM, ETM+, OLI, TIIRS (4,5,7,8,9) (GEE)
Topographic data: JAXA ALOS Global 30m (GEE)
Permafrost: http://data.snap.uaf.edu/data/Base/Other/LandCarbon/Permafrost/
Historic Climate (PRISM): http://data.snap.uaf.edu/data/Base/AK_771m/historical/PRISM/
Future Climate (CCSM and GFDL): http://data.snap.uaf.edu/data/Base/AK_771m/projected/AR5_CMIP5_models/Projected_Monthly_and_Derived_Temperature_Products_771m_CMIP5_AR5/monthly/
Scripts for that preprocessing are found in this folder

```{r}
library(dplyr)
r_files <- list.files(dir, pattern="FULLmosaic-hist-", full.names = TRUE, recursive = TRUE)

#Training Dataframe was prepared in (../scripts/grf_ag_carbon.py)
#these are in the same order as the raster bands
training_df <- read.csv("data/RandomForest_TrainingDATA_04082025.csv") %>% select(-PSP) %>% select(-starts_with("tsf_"))
```

Add lat/lon as bands to the rasters since they are used in model training. Here I generated systematic points within the landscape amounting to ~10% of the entire raster to make predictions for and do change detection. I also make predictions for the raster cells that correspond to actual plot locations to make comparisons of predictions at a plot and regional level.

```{r}
library(terra)

#files
ref <- rast(paste0(dir, "gee/mosaics/landisReg-FULLmosaic-hist-2015.tif"), lyr=33)#elevation
bounds <- vect("F:/randomForest_Landis/input/shp/FullLandscapeV3_082722.shp")
fieldSites <- vect("F:/randomForest_Landis/input/shp/cafiPlotsCropped.shp")

plot(bounds, main = "Field Sites")
plot(fieldSites, add = TRUE, pch = 15, col = "purple")

#sample the landscape (very small proportion of the landscape but should be geographically representative)
regularPoints <- spatSample(ref, size=200000, method = "regular", as.points = TRUE, ext = bounds, na.rm=TRUE)
#randomPoints <- spatSample(ref, size=180000, method = "random", as.points = TRUE, ext = bounds, na.rm=TRUE)

plot(bounds, main = "Points to predict")
plot(sample(regularPoints, 1000), add = TRUE, pch = 15, col = "black")#just plotting 1000
#plot(bounds, main = "Random Points for predictions")
#plot(sample(randomPoints, 1000), add = TRUE, pch = 15, col = adjustcolor("blue", alpha.f = 0.85))
```

This function is set up to read in a raster and remove layers that I duplicated by mistake. It then subsets 200000 pixels from the image using a regular sampling method. It is additionally set up to remove climate layers and retain only 30m layers for higher resolution maps but that is currently commented out.

```{r}

#define function to add xy bands
read_raster <- function (file) {
  r <- rast(file)
  
  #remove bands that I managed to duplicate 
  if (nlyr(r) > 41) r <- terra::subset(r, 1:41)
  
  #filter to our systematic points scheme
  r_vals <- terra::extract(r, regularPoints, xy = TRUE) %>% select(-ID)
  r_vals$dep2_permafrost <- as.integer(r_vals$dep2_permafrost)#dbl to int
  
  #entire image with no climate
  # chunks <- list()
  # total_rows <- nrow(r)
  # total_cols <- ncol(r)
  # chunk_n <- 30
  # rows_per_chunk <- ceiling(total_rows / chunk_n)
  # for (i in 1:chunk_n) {
  #   start_row <- (i - 1) * rows_per_chunk + 1
  #   end_row <- min(i * rows_per_chunk, total_rows)
  #   chunk_raster <- crop(r, ext(0, total_cols, start_row, end_row))
  #   vals <- as.data.frame(chunk_raster, xy=TRUE) ; gc()
  #   clean_df <- vals %>% select(-starts_with(c("tsf", "severity")))
  #   chunks[[i]] <- clean_df
  # }

  
  #something going on with tsf variables (sub r_vals here for predicting through time)
  r_vals <- r_vals %>% select(-starts_with("tsf"))
  # r_vals$tsf_0_10 <- as.character(as.logical(r_vals$tsf_0_10))
  # r_vals$tsf_10_20 <- as.character(as.logical(r_vals$tsf_10_20))
  # r_vals$tsf_20_25 <- as.character(as.logical(r_vals$tsf_20_25))
  # r_vals <- r_vals[complete.cases(r_vals), ]
  return(r_vals)
}
```

Define some functions to make predictions and plot these estimates against our validation set. I'm using gsub and  strsplit to grab unique identifiers from file names so I can store them in a dataframe for later plotting. This requires a final fit from a model workflow. These functions are nested in the model training function.

```{r}
library(data.table)
predict_from_model <- function(file, fit, species){
  
  #naming conventions
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  scenario <- strsplit(basename(file), "-")[[1]][3]
  yr <- gsub(".*-(\\d+)\\.[a-zA-Z]+$", "\\1", basename(file))
  raster <- read_raster(file)


  #prediction
  print(paste0("Processing ", scenario, " scenario ", "for ", spp, " ", "YEAR: ", yr))
  
  #2000000 pixels (I intentionally left this number lower for making predictions through time)
  predicted <- predict(fit, na.omit(raster)) #%>% 
    # summarise(AvgCarbon_gm2=mean(.pred, na.rm = TRUE),
    #           SdCarbon_gm2=sd(.pred, na.rm = TRUE))
  

  #join the predictions to the plot locations and convert to data table
  dt <- setDT(as.data.frame(regularPoints))[, AG_C_gm2 :=rep(predicted$.pred, length.out = nrow(regularPoints))]
  dt <- dt[, -c(1)]
  dt$x <- raster$x
  dt$y <- raster$y
  #add observed CAFI 
  original_obs <- training_df %>% select(Carbon_gm2,x,y) %>% rename(AG_C_gm2=Carbon_gm2)
  dt_full <- rbind(dt, original_obs)
  
  #add in additional info
  dt_full$ClimateScenario <- scenario
  dt_full$Species <- spp
  dt_full$Year <- yr

  
  
  rm(raster, dt)
  return(dt_full)
}

make_validation_df <- function(fit, test_data, species, mean_rmse, mean_rsq){
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  predicted <- predict(fit, test_data)
  observed <- test_data$Carbon_gm2
  vals <- predicted[[1]]
  df <- rbind(data.frame(Observed=observed, Predicted=vals, Species=spp, RMSE=mean_rmse, RSQ=mean_rsq))
  return(df)
}
```

Model training and fitting function. This is the main function where I split my data into a training and validation set, I then run models across different configurations of split data to return a model which has the best RMSE. I then use cross validation again to perform final fits across different configurations of split data to obtain and average RMSE and the best Variance explained (these are the metrics returned for RF, Ranger, and XGBoost)

I additionally add in models for nearest neighbors, support vector machine, and conditional forest but do not fine tune them.

```{r}
library(purrr)
library(tidymodels)
library(randomForest)
library(kknn)
library(kernlab)
#library(partykit)
library(vip)

species <- c("black spruce", "white spruce", "resin birch", "quaking aspen")

train_model <- function(df, species, response="Carbon_gm2") {
  print(paste0("running model for: ", species))
  
  #filter to one species
  training_df_one_spp <- training_df %>% filter(Species == species)
  predictors <- setdiff(names(training_df_one_spp), c("Species", "Carbon_gm2"))

  #train and validation set (stratify the predictor dataset based on nature of the data measured in situ)
  split_data <- initial_split(training_df_one_spp %>% select(all_of(response), all_of(predictors)), prop = 0.95, strata = all_of(response))
  
  #training and test set
  train_data <- training(split_data)
  test_data <- testing(split_data)
  formula <- as.formula(paste(response, "~", paste(predictors, collapse = "+")))
  
  #tidy models recipe (swapping in entire dataset)
  recipe <- recipe(formula = formula, data = train_data) #%>% 
    #step_nzv(all_predictors()) %>%
    #step_corr(all_numeric_predictors(), threshold = 0.80, use = "pairwise.complete.obs", method = "pearson")
  
  #define models to run (nts: cite packages)
  #use additional script to set up cforest to be tidymodels friendly
  #source("scripts/define_cforest_model.R")
  m_rf <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("randomForest")

  
  #params and grids
  rf_grid <- expand.grid(mtry = seq(5, 20, 5), trees = seq(200,500,100), min_n = seq(1, 12, 3))
  
  #cross validation
  nfold <- 5
  print(paste0("performing cross validation across ", nfold, " folds"))
  pv_folds <- vfold_cv(train_data, v = nfold, strata = all_of(response))

  #store the results for all models (selecting the best model across the cross-val folds based on RMSE)
  tuned_results <- workflow() %>% add_recipe(recipe) %>% add_model(m_rf) %>% 
    tune_grid(resamples = pv_folds, grid = rf_grid, metrics = metric_set(rmse, rsq, mae))
  autoplot(tuned_results)
  best_params <- select_best(tuned_results, metric="rmse")
  wflow <- finalize_workflow(workflow() %>% add_recipe(recipe) %>% add_model(m_rf), best_params)
  
  #fit across the resamples
  fit <- fit_resamples(wflow, resamples = pv_folds, metrics = metric_set(rmse, rsq, mae))
  mean_rmse <- collect_metrics(fit) %>% filter(.metric == "rmse") %>% pull(mean)
  mean_rsq <- collect_metrics(fit) %>% filter(.metric == "rsq") %>% pull(mean)
  
  #print(paste0("MODEL: ", model$engine))
  print(paste("Mean RMSE across folds: ", mean_rmse))
  print(paste("Mean R² across folds: ", mean_rsq))
  
  final_fit <- fit(wflow, training_df_one_spp)
  finalized_model <- extract_fit_parsnip(final_fit)
  
  # importance scores (permuation based)
  importance <- vi(finalized_model)
  vip(importance, num_features = 15, geom = "point")
  
  #  return(final_fit)
  #})
  
  #generate predictions (using the regular RF model fit)
  all_predictions <- lapply(unique(r_files), function(file) {predict_from_model(file, final_fit, species)})
  all_predictions_df <- bind_rows(all_predictions)
  
  #generate validation data for plotting
  #validation_fit <- fit(wflow, test_data)
  #validation_pred_df <- make_validation_df(validation_fit, test_data, species, mean_rmse, mean_rsq)
    
  return(list(train_dfs=train_data, test_dfs=test_data, final_models=finalized_model, 
              all_predictions=all_predictions_df, validation_data=validation_pred_df))
}

#run the models
model_output <- lapply(species, function(spp){train_model(training_df,species=spp)})
predictions <- do.call(rbind, lapply(model_output[1:4], function(x) x$all_predictions))
validations <- do.call(rbind, lapply(model_output[1:4], function(x) x$validation_data))
#write.csv(predictions, "F:/jetstream/data/PredictedCarbon_RasterValues.csv", row.names = FALSE)
```

Graph the predicted vs observed data for the validation fits.

```{r}
library(ggplot2)
library(viridis)

#plotting validation data vs predictions for each species with a 60/40 split
validation_plot <- ggplot(validations, aes(x = Observed, y = Predicted, color = abs(Predicted-Observed))) +
  #geom_smooth(method = "lm", color = "black", linetype = "dashed", se = FALSE)+
  geom_point(size = 5, alpha = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  scale_color_viridis(option = "H", direction=-1)+
  facet_grid(~ Species, nrow(1)) +
  theme_bw() +
  scale_x_continuous(limits=c(0,500))+
  scale_y_continuous(limits=c(0,500))+
  labs(x = "Observed", y = "Predicted") +
  theme(legend.position = "none", strip.text = element_text(size = 14), axis.title = element_text(size = 16)) +
  geom_text(aes(x = 0.9 * max(Observed), y = 0.1 * max(Predicted), label = paste("r2:", round(RSQ, 2), "\nRMSE:", round(RMSE, 2))),
            size = 10, hjust = 1, vjust = 0, color = "black")

plot(validation_plot)
```

Graph carbon through time from the rasters we made predictions on.

```{r}

```

After looking at an initial carbon plot through time I decided to build a new dataset from the black spruce predictions. Interested to see if predictions will diverge if trained off of more data points and with different climate sources to predict from.

```{r}
years <- seq(2006, 2024)
vars <- c("pr", "tas")
scenarios <- c("CCSM4_rcp85", "GFDL-CM3_rcp85")
dir <- "F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/ClimateData/"
pattern <- paste0("^(", paste(vars, collapse = "|"), ").*(", 
                  paste(scenarios, collapse = "|"), ").*(", 
                  paste(years, collapse = "|"), ")\\.tif$")

climate_files <- list.files(dir, pattern = pattern, full.names = TRUE, recursive = TRUE)

#grab predictions from one species
predictions_Bspruce <- predictions %>% filter(ClimateScenario == "hist", Species == "Black spruce")

# Read in the climate data
subset_clim_data <- function(file){
  r <- rast(file)
  
  bname <- basename(file)
  var <- strsplit(bname, "_")[[1]][1]
  scenario <- strsplit(bname, "_")[[1]][5]
  pway <- strsplit(bname, "_")[[1]][6]
  
  yr <- gsub(".*_(\\d{4}).*", "\\1", bname)
  month <- gsub(".*_(\\d{2})_.*", "\\1", bname)
  #clip
  #crs(r) <- crs(ref)
  r_resampled <- project(r, ref)
  clipped <- mask(r_resampled, bounds)
  #res(clipped) <- res(r)
  names(clipped) <- paste0(var, month)
  layer_name <-  paste0(var, month)
  
  #extract a subset of the raster values
  r_vals <- terra::extract(clipped, regularPoints, xy = TRUE) %>% select(-ID)
  fname <- paste0("Subset_", bname)
  
  regularPoints[[paste0(layer_name)]] <- r_vals
  smaller_r <- rasterize(regularPoints, clipped, field = layer_name)
  names(smaller_r) <- layer_name
  smaller_r$scenario <- paste0(scenario, pway)
  smaller_r$x <- r_vals$x
  smaller_r$y <- r_vals$y
  
  #writeRaster(smaller_r, filename = fname, overwrite = TRUE)
  rm(r)
  return(smaller_r)
}

#apply the function then merge the datasets
processed_gfdl_pr <- lapply(climate_files[grepl("pr_total_mm_AR5_GFDL", climate_files)], subset_clim_data)
processed_ccsm_pr <- lapply(climate_files[grepl("pr_total_mm_AR5_CCSM", climate_files)], subset_clim_data)
processed_gfdl_tas <- lapply(climate_files[grepl("tas_mean_C_AK_AR5_GFDL", climate_files)], subset_clim_data)
processed_ccsm_tas <- lapply(climate_files[grepl("tas_mean_C_AK_AR5_CCSM", climate_files)], subset_clim_data)

gfdl_pr <- bind_rows(processed_gfdl_pr)
ccsm_pr <- bind_rows(processed_ccsm_pr)
gfdl_tas <- bind_rows(processed_gfdl_tas)
ccsm_tas <- bind_rows(processed_ccsm_tas)
```
