---
title: "Random Forest Mapping"
author: "Wesley Rancher"
date: "2025-04-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# title: Random Forest Modeling
# author: Wesley Rancher
# date: "2025-1-22"

setwd("F:/jetstream/")
dir <- "F:/remoteSensing/data/output/"
```

Read in training dataset and rasters to make predictions on

Landsat data MSS, TM, ETM+, OLI, TIIRS (4,5,7,8,9) (GEE)
Topographic data: JAXA ALOS Global 30m (GEE)
Permafrost: http://data.snap.uaf.edu/data/Base/Other/LandCarbon/Permafrost/
Historic Climate (PRISM): http://data.snap.uaf.edu/data/Base/AK_771m/historical/PRISM/
Future Climate (CCSM and GFDL): http://data.snap.uaf.edu/data/Base/AK_771m/projected/AR5_CMIP5_models/Projected_Monthly_and_Derived_Temperature_Products_771m_CMIP5_AR5/monthly/
Scripts for that preprocessing are found in this folder

```{r}
library(dplyr)
training_df <- read.csv("../../randomForest_Landis/output/main_predictor_data.csv") %>% select(-PSP, -scenario)
```

Define some functions to make predictions and plot these estimates against our validation set. I'm using gsub and  strsplit to grab unique identifiers from file names so I can store them in a dataframe for later plotting. This requires a final fit from a model workflow. These functions are nested in the model training function.

```{r}
library(data.table)
pfun <- function(object, new_data){
  predicted <- predict(object, new_data)$predictions
}

make_validation_df <- function(fit, test_data, species, mean_rmse, mean_rsq){
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  predicted <- predict(fit, test_data)
  observed <- test_data$ag_carbon
  vals <- predicted[[1]]
  df <- rbind(data.frame(Observed=observed, Predicted=vals, Species=spp, RMSE=mean_rmse, RSQ=mean_rsq))
  return(df)
}
```

Model training and fitting function. This is the main function where I split my data into a training and validation set, I then run models across different configurations of split data to return a model which has the best RMSE. I then use cross validation again to perform final fits across different configurations of split data to obtain and average RMSE and the best Variance explained (these are the metrics returned for RF, Ranger, and XGBoost)

I additionally add in models for nearest neighbors, support vector machine, and conditional forest but do not fine tune them.

```{r}
library(purrr)
library(tidymodels)
library(randomForest)
library(kknn)
library(kernlab)
#library(partykit)
library(vip)

species <- c("BlackSpruce", "WhiteSpruce", "PaperBirch", "QuakingAspen")

train_model <- function(df, species, response="ag_carbon") {
  print(paste0("running model for: ", species))
  
  #filter to one species
  training_df_one_spp <- training_df %>% filter(spp == species)
  predictors <- setdiff(names(training_df_one_spp), c("spp", "ag_carbon"))

  #train and validation set (stratify the predictor dataset based on nature of the data measured in situ)
  split_data <- initial_split(training_df_one_spp %>% select(all_of(response), all_of(predictors)), prop = 0.80, strata = all_of(response))
  
  #training and test set
  train_data <- training(split_data)
  test_data <- testing(split_data)
  formula <- as.formula(paste(response, "~", paste(predictors, collapse = "+")))
  
  #tidy models recipe (swapping in entire dataset)
  recipe <- recipe(formula = formula, data = train_data) %>% 
    step_nzv(all_predictors()) %>%
    step_corr(all_numeric_predictors(), threshold = 0.80, use = "pairwise.complete.obs", method = "pearson")
  
  #define models to run (nts: cite packages)
  #use additional script to set up cforest to be tidymodels friendly
  #source("scripts/define_cforest_model.R")
  m_rf <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("randomForest")
  m_ranger <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("ranger")
  m_xgb <- boost_tree(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("xgboost")
  m_kknn <- nearest_neighbor(neighbors=tune(), weight_func = tune(), dist_power = tune()) %>% set_mode("regression") %>% set_engine("kknn")
  m_svm <- svm_rbf(cost= tune(), rbf_sigma = tune()) %>% set_mode("regression") %>% set_engine("kernlab")
  #m_cforest <- cforest() %>% set_mode("regression") %>% set_engine("partykit")
  models <- list(rf=m_rf,ranger=m_ranger)
  
  #params and grids
  #top3_params <- parameters(mtry(), trees(), min_n())
  # I think these two need recipes where the data is normalized
  kknn_params <- parameters(neighbors(range = c(4, 25)), weight_func(values = c("rectangular", "triangular", "biweight", 
                                                                                "triweight", "epanechnikov", "gaussian")),dist_power(range = c(1, 2)))  
  svm_params <- parameters(cost(range = c(-8, 4), trans = scales::log10_trans()), rbf_sigma(range = c(-3, 0), trans = scales::log10_trans()))
    
  #rf, ranger, and xgb use the same params
  rf_grid <- expand.grid(mtry = seq(5, 20, 5), trees = seq(200,500,100), min_n = seq(1, 12, 3))

  kknn_grid <- grid_regular(kknn_params, levels = c(4,3,2))
  svm_grid <- grid_regular(svm_params, levels = 10)
  grids <- list(rf_grid=rf_grid, ranger_grid=rf_grid)
  
  #cross validation
  nfold <- 5
  print(paste0("performing cross validation across ", nfold, " folds"))
  pv_folds <- vfold_cv(train_data, v = nfold, strata = all_of(response))

  #store the results for all models (selecting the best model across the cross-val folds based on RMSE)
  #working outside of the map2 function to generate predictions
  
  #results <- map2(models, grids, function(model, grid) {
  #trying to leave this dynamic
  tuned_results <- workflow() %>% add_recipe(recipe) %>% add_model(m_rf) %>% 
    tune_grid(resamples = pv_folds, grid = rf_grid, metrics = metric_set(rmse, rsq, mae))
  autoplot(tuned_results)
  best_params <- select_best(tuned_results, metric="rmse")
  wflow <- finalize_workflow(workflow() %>% add_recipe(recipe) %>% add_model(m_rf), best_params)
  
  #fit across the resamples
  fit <- fit_resamples(wflow, resamples = pv_folds, metrics = metric_set(rmse, rsq, mae))
  mean_rmse <- collect_metrics(fit) %>% filter(.metric == "rmse") %>% pull(mean)
  mean_rsq <- collect_metrics(fit) %>% filter(.metric == "rsq") %>% pull(mean)
  
  #print(paste0("MODEL: ", model$engine))
  print(paste("Mean RMSE across folds: ", mean_rmse))
  print(paste("Mean R² across folds: ", mean_rsq))
  
  final_fit <- fit(wflow, training_df_one_spp)
  finalized_model <- extract_fit_parsnip(final_fit)
  
  # importance scores (permuation based)
  importance <- vi(finalized_model)
  vip(importance, num_features = 15, geom = "point")
  
  #  return(final_fit)
  #})
  
  #generate predictions (using the regular RF model fit)
  #all_predictions <- lapply(unique(r_files), function(file) {predict_from_model(file, final_fit, species)})
  #all_predictions_df <- bind_rows(all_predictions)
  
  #generate validation data for plotting
  validation_fit <- fit(wflow, test_data)
  validation_pred_df <- make_validation_df(validation_fit, test_data, species, mean_rmse, mean_rsq)
    
  return(list(train_dfs=train_data, test_dfs=test_data, final_models=finalized_model, 
              all_predictions=all_predictions_df, validation_data=validation_pred_df))
}

#run the models
model_output <- lapply(species, function(spp){train_model(training_df,species=spp)})
predictions <- do.call(rbind, lapply(model_output[1:4], function(x) x$all_predictions))
validations <- do.call(rbind, lapply(model_output[1:4], function(x) x$validation_data))
#write.csv(predictions, "F:/jetstream/data/PredictedCarbon_RasterValues.csv", row.names = FALSE)
```

Graph the predicted vs observed data for the validation fits.

```{r}
library(ggplot2)
library(viridis)

#plotting validation data vs predictions for each species with a 60/40 split
validation_plot <- ggplot(validations, aes(x = Observed, y = Predicted, color = abs(Predicted-Observed))) +
  #geom_smooth(method = "lm", color = "black", linetype = "dashed", se = FALSE)+
  geom_point(size = 5, alpha = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  scale_color_viridis(option = "H", direction=-1)+
  facet_grid(~ Species, nrow(1)) +
  theme_bw() +
  scale_x_continuous(limits=c(0,500))+
  scale_y_continuous(limits=c(0,500))+
  labs(x = "Observed", y = "Predicted") +
  theme(legend.position = "none", strip.text = element_text(size = 14), axis.title = element_text(size = 16)) +
  geom_text(aes(x = 0.9 * max(Observed), y = 0.1 * max(Predicted), label = paste("r2:", round(RSQ, 2), "\nRMSE:", round(RMSE, 2))),
            size = 10, hjust = 1, vjust = 0, color = "black")

plot(validation_plot)
```

Graph carbon through time from the rasters we made predictions on.

```{r}

```

After looking at an initial carbon plot through time I decided to build a new dataset from the black spruce predictions. Interested to see if predictions will diverge if trained off of more data points and with an additional stream.

```{r}
years <- seq(2006, 2024)
dir <- "F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/ClimateData/"
climate_files <- list.files(dir, pattern = paste0("^(tas.*(", paste(years, collapse = "|"), "))\\.tif$"),full.names = TRUE,recursive = TRUE)#2006-2024

#grab predictions from one species
predictions_Bspruce <- predictions %>% filter(ClimateScenario == "hist", Species == "Black spruce")

# Read in the climate data
join_climate_to_predictions <- function(file){
  r <- rast(file)
  
  bname <- basename(file)
  var <- strsplit(bname, "_")[[1]][1]
  scenario <- strsplit(bname, "_")[[1]][5]
  pway <- strsplit(bname, "_")[[1]][6]
  
  yr <- gsub(".*_(\\d{4}).*", "\\1", bname)
  month <- gsub(".*_(\\d{2})_.*", "\\1", bname)
  #clip
  #crs(r) <- crs(ref)
  r_resampled <- project(r, ref)
  clipped <- mask(r_resampled, bounds)
  res(clipped) <- res(r)
  names(clipped) <- paste0(var, month)
  layer_name <-  paste0(var, month)
  
  #extract a subset of the raster values
  r_vals <- terra::extract(clipped, regularPoints, xy = TRUE) %>% select(-ID)
  fname <- paste0("Subset_", bname)
  
  regularPoints[[paste0(layer_name)]] <- r_vals
  smaller_r <- rasterize(regularPoints, clipped, field = layer_name)
  names(smaller_r) <- layer_name
  writeRaster(smaller_r, filename = fname, overwrite = TRUE)
  
  #now we treat the predicted as our "observed"
  joined_vals <- merge(r_vals, predictions_Bspruce, by = c("x","y"))%>%
    select(-paste0(layer_name,".x"))%>%
    rename(paste0(layer_name) == paste0(layer_name, ".y"))
  
  #join
  r_vals$prediction <- predictions_Bspruce$prediction

                       
  return(joined_vals)
}


joined_predictions_new_climate <- lapply(unique(climate_files), function(climate_files) {join_climate_to_predictions(file, final_fit, species)})


```
