---
title: "Random Forest Mapping"
author: "Wesley Rancher"
date: "2025-04-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# title: Random Forest Modeling
# author: Wesley Rancher
# date: "2025-1-22"

setwd("F:/jetstream/")
dir <- "F:/remoteSensing/data/output/"
```

Read in training dataset and rasters to make predictions on

Landsat data MSS, TM, ETM+, OLI, TIIRS (4,5,7,8,9) (GEE)
Topographic data: JAXA ALOS Global 30m (GEE)
Permafrost: http://data.snap.uaf.edu/data/Base/Other/LandCarbon/Permafrost/
Historic Climate (PRISM): http://data.snap.uaf.edu/data/Base/AK_771m/historical/PRISM/
Future Climate (CCSM and GFDL): http://data.snap.uaf.edu/data/Base/AK_771m/projected/AR5_CMIP5_models/Projected_Monthly_and_Derived_Temperature_Products_771m_CMIP5_AR5/monthly/
Scripts for that preprocessing are found in this folder

```{r}
library(dplyr)
r_files <- list.files(dir, pattern="FULLmosaic-", full.names = TRUE, recursive = TRUE)

#Training Dataframe was prepared in (../scripts/grf_ag_carbon.py)
#these are in the same order as the raster bands
training_df <- read.csv("data/RandomForest_TrainingDATA_04082025.csv") %>% select(-PSP)
```

Add lat/lon as bands to the rasters since they are used in model training. Here I generated systematic points within the landscape amounting to ~10% of the entire raster to make predictions for and do change detection. I also make predictions for the raster cells that correspond to actual plot locations to make comparisons of predictions at a plot and regional level.

```{r}
library(terra)

#files
ref <- rast(paste0(dir, "gee/mosaics/landisReg-FULLmosaic-hist-2015.tif"), lyr=33)#elevation
bounds <- vect("F:/randomForest_Landis/input/shp/FullLandscapeV3_082722.shp")
fieldSites <- vect("F:/randomForest_Landis/input/shp/cafiPlotsCropped.shp")

plot(bounds, main = "Field Site Locations")
plot(fieldSites, add = TRUE, pch = 15, col = "purple")

#sample the landscape (very small proportion of the landscape but should be geographically representative)
regularPoints <- spatSample(ref, size=200000, method = "regular", as.points = TRUE, ext = bounds, na.rm=TRUE)
#randomPoints <- spatSample(ref, size=180000, method = "random", as.points = TRUE, ext = bounds, na.rm=TRUE)

plot(bounds, main = "Regular Points for predictions")
plot(sample(regularPoints, 1000), add = TRUE, pch = 15, col = "deeppink2")#just plotting 1000
#plot(bounds, main = "Random Points for predictions")
#plot(sample(randomPoints, 1000), add = TRUE, pch = 15, col = adjustcolor("blue", alpha.f = 0.85))
```

```{r}
#define function to add xy bands
read_raster <- function (file) {
  r <- rast(file)
  
  #remove bands that I managed to duplicate 
  if (nlyr(r) > 41) r <- terra::subset(r, 1:41)
  
  #filter to our systematic points scheme
  r_vals <- terra::extract(r, regularPoints, xy = TRUE) %>% select(-ID)
  r_vals$dep2_permafrost <- as.integer(r_vals$dep2_permafrost)#dbl to int
  return(r_vals)
}
```

Define some functions to make predictions and plot these estimates against our validation set. I'm using gsub and  strsplit to grab unique identifiers from file names so I can store them in a dataframe for later plotting. This requires a final fit from a model workflow. These functions are nested in the model training function.

```{r}
pfun <- function(object, new_data){
  predicted <- predict(object, new_data)$predictions
}
  
predict_from_model <- function(file, fit, species){
  
  #naming conventions
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  scenario <- strsplit(basename(file), "-")[[1]][3]
  yr <- gsub(".*-(\\d+)\\.[a-zA-Z]+$", "\\1", basename(file))
  raster <- read_raster(file)
  
  #prediction
  print(paste0("Processing ", scenario, " scenario ", "for ", spp, " ", "YEAR: ", yr))
  
  #2000000 pixels (I intentionally left this number lower for making predictions through time)
  predicted <- predict(fit, na.omit(raster)) %>% 
    summarise(AvgCarbon_gm2=mean(.pred, na.rm = TRUE),
              SdCarbon_gm2=sd(.pred, na.rm = TRUE))
  
  predicted$ClimateScenario <- scenario
  predicted$Species <- spp
  predicted$Year <- yr
  
  rm(raster)
  return(as.data.frame(predicted))
}

make_validation_df <- function(fit, test_data, species, mean_rmse, mean_rsq){
  spp <- tools::toTitleCase(strsplit(species, " ")[[1]][1])%>%paste0(" ", strsplit(species, " ")[[1]][2])
  predicted <- predict(fit, test_data)
  observed <- test_data$Carbon_gm2
  vals <- predicted[[1]]
  df <- rbind(data.frame(Observed=observed, Predicted=vals, Species=spp, RMSE=mean_rmse, RSQ=mean_rsq))
  return(df)
}
```

Model training and fitting function. This is the main function where I split my data into a training and validation set, I then run models across different configurations of split data to return a model which has the best RMSE. I then use cross validation again to perform final fits across different configurations of split data to obtain and average RMSE and the best Variance explained (these are the metrics returned for RF, Ranger, and XGBoost)

I additionally add in models for nearest neighbors, support vector machine, and conditional forest but do not fine tune them.

```{r}
library(purrr)
library(tidymodels)
library(randomForest)
library(kknn)
library(kernlab)
#library(partykit)
library(vip)

species <- c("black spruce", "white spruce", "resin birch", "quaking aspen")

train_model <- function(df, species, response="Carbon_gm2") {
  print(paste0("running model for: ", species))
  
  #filter to one species
  training_df_one_spp <- training_df %>% filter(Species == species)
  predictors <- setdiff(names(training_df_one_spp), c("Species", "Carbon_gm2"))

  #train and validation set (stratify the predictor dataset based on nature of the data measured in situ)
  split_data <- initial_split(training_df_one_spp %>% select(all_of(response), all_of(predictors)), prop = 0.75, strata = all_of(response))
  
  #training and test set
  train_data <- training(split_data)
  test_data <- testing(split_data)
  formula <- as.formula(paste(response, "~", paste(predictors, collapse = "+")))
  
  #tidy models recipe
  recipe <- recipe(formula = formula, data = train_data) #%>% 
    #step_nzv(all_predictors()) %>%
    #step_corr(all_numeric_predictors(), threshold = 0.80, use = "pairwise.complete.obs", method = "pearson")
  
  #define models to run (nts: cite packages)
  #use additional script to set up cforest to be tidymodels friendly
  #source("scripts/define_cforest_model.R")
  m_rf <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("randomForest")
  m_ranger <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("ranger")
  m_xgb <- boost_tree(mtry = tune(), trees = tune(), min_n = tune()) %>% set_mode("regression") %>% set_engine("xgboost")
  m_kknn <- nearest_neighbor(neighbors=tune(), weight_func = tune(), dist_power = tune()) %>% set_mode("regression") %>% set_engine("kknn")
  m_svm <- svm_rbf(cost= tune(), rbf_sigma = tune()) %>% set_mode("regression") %>% set_engine("kernlab")
  #m_cforest <- cforest() %>% set_mode("regression") %>% set_engine("partykit")
  models <- list(rf=m_rf,ranger=m_ranger)
  
  #params and grids
  #top3_params <- parameters(mtry(), trees(), min_n())
  # I think these two need recipes where the data is normalized
  kknn_params <- parameters(neighbors(range = c(4, 25)), weight_func(values = c("rectangular", "triangular", "biweight", 
                                                                                "triweight", "epanechnikov", "gaussian")),dist_power(range = c(1, 2)))  
  svm_params <- parameters(cost(range = c(-8, 4), trans = scales::log10_trans()), rbf_sigma(range = c(-3, 0), trans = scales::log10_trans()))
    
  #rf, ranger, and xgb use the same params
  if (species == "black spruce"){
    rf_grid <- expand.grid(mtry = seq(5, 20, 5), trees = seq(300,500,25), min_n = seq(1, 12, 2))
  } else if (species == "white spruce") {
    rf_grid <- expand.grid(mtry = seq(5, 20, 3), trees = seq(200,1000,100), min_n = seq(1, 15, 3))
  } else if (species == "resin birch"){
    rf_grid <- expand.grid(mtry = seq(5, 20, 5), trees = seq(300,500,25), min_n = seq(1, 12, 2))
  } else 
    rf_grid <- expand.grid(mtry = seq(5, 20, 3), trees = seq(200,1000,100), min_n = seq(1, 15, 3))

  kknn_grid <- grid_regular(kknn_params, levels = c(4,3,2))
  svm_grid <- grid_regular(svm_params, levels = 10)
  grids <- list(rf_grid=top3_grid, ranger_grid=top3_grid)
  
  #cross validation
  nfold <- 12
  print(paste0("performing cross validation across ", nfold, " folds"))
  pv_folds <- vfold_cv(train_data, v = nfold, strata = all_of(response))

  #store the results for all models (selecting the best model across the cross-val folds based on RMSE)
  #working outside of the map2 function to generate predictions
  
  #results <- map2(models, grids, function(model, grid) {
  #trying to leave this dynamic
  tuned_results <- workflow() %>% add_recipe(recipe) %>% add_model(m_rf) %>% 
    tune_grid(resamples = pv_folds, grid = top3_grid, metrics = metric_set(rmse, rsq, mae))
  autoplot(tuned_results)
  best_params <- select_best(tuned_results, metric="rmse")
  wflow <- finalize_workflow(workflow() %>% add_recipe(recipe) %>% add_model(m_rf), best_params)
  
  #fit across the resamples
  fit <- fit_resamples(wflow, resamples = pv_folds, metrics = metric_set(rmse, rsq, mae))
  mean_rmse <- collect_metrics(fit) %>% filter(.metric == "rmse") %>% pull(mean)
  mean_rsq <- collect_metrics(fit) %>% filter(.metric == "rsq") %>% pull(mean)
  
  #print(paste0("MODEL: ", model$engine))
  print(paste("Mean RMSE across folds: ", mean_rmse))
  print(paste("Mean R² across folds: ", mean_rsq))
  
  final_fit <- fit(wflow, train_data)
  finalized_model <- extract_fit_parsnip(final_fit)
  
  # importance scores (permuation based)
  importance <- vi(finalized_model)
  vip(importance, num_features = 15, geom = "point")
  #  return(final_fit)
  #})
  
  #generate predictions (using the regular RF model fit)
  all_predictions <- lapply(unique(r_files), function(file) {predict_from_model(file, final_fit, species)})
  all_predictions_df <- bind_rows(all_predictions)
  
  #generate validation data for plotting
  validation_fit <- fit(wflow, test_data)
  validation_pred_df <- make_validation_df(validation_fit, test_data, species, mean_rmse, mean_rsq)
    
  return(list(train_dfs=train_data, test_dfs=test_data, final_models=finalized_model, 
              all_predictions=all_predictions_df, validation_data=validation_pred_df))
}

#run the models
model_output <- lapply(species, function(spp){train_model(training_df,species=spp)})
predictions <- do.call(rbind, lapply(model_output[1:4], function(x) x$all_predictions))
validations <- do.call(rbind, lapply(model_output[1:4], function(x) x$validation_data))
write.csv(predictions, "F:/jetstream/data/PredictedCarbon_RasterValues.csv", row.names = FALSE)
```

Graph the predicted vs observed data for the validation fits.

```{r}
library(ggplot2)
library(viridis)

#plotting validation data vs predictions for each species with a 60/40 split
validation_plot <- ggplot(validations, aes(x = Observed, y = Predicted, color = abs(Predicted-Observed))) +
  #geom_smooth(method = "lm", color = "black", linetype = "dashed", se = FALSE)+
  geom_point(size = 5, alpha = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  scale_color_viridis(option = "H", direction=-1)+
  facet_grid(~ Species, nrow(1)) +
  theme_bw() +
  scale_x_continuous(limits=c(0,500))+
  scale_y_continuous(limits=c(0,500))+
  labs(x = "Observed", y = "Predicted") +
  theme(legend.position = "none", strip.text = element_text(size = 14), axis.title = element_text(size = 16)) +
  geom_text(aes(x = 0.9 * max(Observed), y = 0.1 * max(Predicted), label = paste("r2:", round(RSQ, 2), "\nRMSE:", round(RMSE, 2))),
            size = 10, hjust = 1, vjust = 0, color = "black")

plot(validation_plot)
```

Graph carbon through time from the rasters we made predictions on.

```{python}
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns

predicted_data = pd.read_csv("data/PredictedCarbon_RasterValues.csv")

predictions['Year'] = predictions['Year'].astype(int)

# Set a style
sns.set(style="whitegrid")

# Get unique species for subplots
species_list = predictions['Species'].unique()
n_species = len(species_list)

# Create subplots
fig, axes = plt.subplots(n_species, 1, figsize=(10, 5 * n_species), sharex=True)

if n_species == 1:
    axes = [axes]  # Ensure axes is iterable

# Plot for each species
for ax, species in zip(axes, species_list):
    data = predictions[predictions['Species'] == species]
    sns.lineplot(data=data, x='Year', y='AvgCarbon_gm2', hue='ClimateScenario', ax=ax, marker='o')
    ax.set_title(species)
    ax.set_ylabel("AvgCarbon (g/m²)")
    ax.legend(title='Climate Scenario')

axes[-1].set_xlabel("Year")
plt.tight_layout()
plt.show()
```