---
title: "Supplemental Figures"
author: "Wesley Rancher"
date: "2025-04-10"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
dir <- "F:/jetstream/data/"
setwd(dir)
library(terra)
library(dplyr)
library(ggplot2)
```

# Plotting Landis Climate Data and Remote Sensing Climate Data

I obtained climate data from:

Historic Climate (PRISM): http://data.snap.uaf.edu/data/Base/AK_771m/historical/PRISM/
Future Climate Temperature (CCSM and GFDL): http://data.snap.uaf.edu/data/Base/AK_771m/projected/AR5_CMIP5_models/Projected_Monthly_and_Derived_Temperature_Products_771m_CMIP5_AR5/monthly/

```{r, Plotting Climate Data, include=FALSE, eval=FALSE}
climatedir <- "F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/ClimateData/"

#clean landis climate monthly log
climateFuture <- read.csv("F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/rep1_4.3.2025/interior_hist_ncar/DGS-succession-monthly-log.csv")
climateFuture_temp <- climateFuture %>%
  select(Time, Ppt, Airtemp, ClimateRegionName)%>%
  mutate(Year=Time+2000)%>%
  group_by(Year, ClimateRegionName)%>%
  summarise(Temp_Avg=mean(Airtemp,na.rm=TRUE),
            Temp_StDev=sd(Airtemp, na.rm = TRUE),
            PR_Avg=mean(Ppt,na.rm=TRUE),
            PR_StDev=sd(Ppt, na.rm = TRUE))


#plotting climate used in RS random forest
years <- seq(2006,2050)
climate_files_snaps <- list.files(climatedir, pattern = paste0("^(tas.*(", paste(years, collapse = "|"), "))\\.tif$"),full.names = TRUE,recursive = TRUE)

#plot by ecoregions
bounds <- vect("F:/Rancher_Sims_Full_Landscape/input/FullLandscapeV3_082722.shp")
ref <- rast("F:/Rancher_Sims_Full_Landscape/input/AK_ClimateMap_10_Regions.tif")
ref_projected <- project(ref, "EPSG:3338")
ref_clipped <- mask(ref_projected, bounds)
randomPoints <- spatSample(ref_projected, size=2000000, xy=TRUE,method = "random", ext = bounds, na.rm=TRUE)


get_average <- function(file){
  r <- rast(file)
  
  bname <- basename(file)
  var <- strsplit(bname, "_")[[1]][1]
  scenario <- strsplit(bname, "_")[[1]][5]
  pway <- strsplit(bname, "_")[[1]][6]
  
  yr <- gsub(".*_(\\d{4}).*", "\\1", bname)
  month <- gsub(".*_(\\d{2})_.*", "\\1", bname)
  #clip
  #crs(r) <- crs(ref)
  r_resampled <- project(r, ref_projected)
  clipped <- mask(r_resampled, bounds)
  #res(clipped) <- res(r)
  names(clipped) <- paste0(var, month)
  layer_name <-  paste0(var, month)
  df <- as.data.frame(clipped, xy=TRUE)
  joined <- inner_join(df, randomPoints, by=c("x", "y"))%>%slice_sample(n=1000)
  joined$var <- layer_name
  colnames(joined)[3]<-var
  joined$Year <- yr
  joined$Month <- month
  joined$ClimateScenario <- paste0(scenario,pway)
  
  print(paste0("processed: ", bname))
  return(joined)
}

temperature_dataframes_list <- lapply(climate_files_snaps, get_average)
temperature_df <- bind_rows(temperature_dataframes_list)

#seasonal line plot (faceted by season lines for )
summary_projected_climate <- temperature_df%>%
  rename(EcoRegions = AK_ClimateMap_10_Regions) %>%
  mutate(Season = case_when(
    Month %in% 3:5 ~ "Spring",
    Month %in% 6:8 ~ "Summer",
    Month %in% 9:11 ~ "Fall",
    TRUE ~ NA_character_)) %>%
  filter(!is.na(Season)) %>%
  group_by(EcoRegions, Year, ClimateScenario, Season) %>%
  summarise(MeanTemp = mean(Temp, na.rm = TRUE), .groups = "drop")

```

## Plotting Fire in LANDIS against Empirical (Directly from Zachary Robbin's tutorial swapped for AK data)

citations: https://github.com/LANDIS-II-Foundation/Extension-Social-Climate-Fire/tree/master/Supporting%20R%20Code/Ignition_FWI


Here we can see the number of lightning ignited fires per year. Comparing landis output log files to data from Karen short:
https://www.fs.usda.gov/rds/archive/Catalog/RDS-2013-0009.4/

This all needs to be updated with results from newer sims

```{r, Short Data}
library(RODBC) 

short_data <- odbcConnectAccess2007("F:/Rancher_Sims_Full_Landscape/FinalSimulationsThesis/Karen_Short_Data/FPA_FOD_20170508.accdb") 
Fires <- sqlFetch(short_data, "Fires")
Fires_AK <- Fires %>%
  filter(STATE=="AK")%>% 
  select(FPA_ID, FIRE_YEAR, DISCOVERY_DOY, STAT_CAUSE_DESCR,CONT_DATE, FIRE_SIZE, LATITUDE, LONGITUDE)

l_fire_days_sort <- Fires_AK %>% filter(STAT_CAUSE_DESCR=="Lightning")
h_fire_dat  <- Fires_AK %>% filter(!STAT_CAUSE_DESCR %in% c("Lightning","Missing/Undefined"))
```

```{r, Empirical # Fires}
library(RColorBrewer)
library(pscl)

#plot no of fires/yr
l_fire_days_sort_count<- l_fire_days_sort%>%
  dplyr::rename(J_DAY=DISCOVERY_DOY)%>%
  group_by(FIRE_YEAR)%>%
  dplyr::summarise(COUNT=n())
#aesthetics
reds <- colorRampPalette(brewer.pal(9, "Oranges"))(length(l_fire_days_sort_count$COUNT))
sorted_counts <- sort(l_fire_days_sort_count$COUNT, decreasing = FALSE)
color_map <- setNames(reds, sorted_counts)
bar_colors <- color_map[as.character(l_fire_days_sort_count$COUNT)]

colnames(l_fire_days_sort_count) <- c("YEAR", "COUNT")

barplot(l_fire_days_sort_count$COUNT, main ="No of ign/yr Lightning Observed",col=bar_colors,names.arg=l_fire_days_sort_count$YEAR)
```

Human ignitions

```{r Human or Accidental fires}
h_fire_days_sort <- h_fire_dat %>%
  dplyr::rename(J_DAY = DISCOVERY_DOY) %>%
  group_by(FIRE_YEAR) %>%
  dplyr::summarise(COUNT = n())
colnames(h_fire_days_sort) <- c("YEAR", "COUNT")

sorted_counts <- sort(h_fire_days_sort$COUNT, decreasing = FALSE)
color_map <- setNames(reds, sorted_counts)
bar_colors <- color_map[as.character(h_fire_days_sort$COUNT)]

barplot(h_fire_days_sort$COUNT, main ="No of ign/yr Human Accidental Observed",col=bar_colors,names.arg=h_fire_days_sort$YEAR)
```

Here we match them up with Fire weather index(FWI) data generated from the historic record of weather fed through LANDIS-II Fire weather to get the models calculation of FWI. See the Climate library documentation for assistance getting FWI.

```{r FWI Relationship to Empirical}
library(plyr)
ign_types <- c("Lightning", "HumanAccidental")

fire_days_list <- list(l_fire_days_sort, h_fire_dat) ##organizing all ignition types into a list


##Import daily historic FWI data
FWI_dat <- read.csv("F:/Rancher_Sims_Full_Landscape/interior_hist_ncar_3.18.2025/Climate-future-daily-input-log.csv")
FWI_dat<-with(FWI_dat,aggregate(FireWeatherindex,by=list(Year=Year,Timestep=Day),FUN=mean))



FWI_dat <- FWI_dat[,c(1,2,3)]
colnames(FWI_dat)[3]<-"FWI"
FWI_dat$ID <- paste(FWI_dat$Year+1999, "_", FWI_dat$Timestep, sep="") #creating date identifier out of date and julian day
igns_list <- list()

for (i in 1:length(ign_types[1:2])){#THIS DOESN'T INCLUDE RX BURNS BUT THATS CAUSE WE ARE PROVIDING THOSE TO SCRAPPLE DIRECT
  ign_type_select <- fire_days_list[[i]] %>%dplyr::rename(YEAR=FIRE_YEAR, J_DAY=DISCOVERY_DOY)##selecting each ignition type individually
  fire_days_count <- ddply(ign_type_select, .(ign_type_select$YEAR, ign_type_select$J_DAY), nrow) #finds duplicate rows in fire data
  colnames(fire_days_count) <- c("YEAR", "JD", "No_FIRES") #Renaming columns for consistency
  fire_days_count$ID <- paste(fire_days_count$YEAR, "_", fire_days_count$JD, sep="") #creating date identifier out of date and julian day

  ##Merging dataframes by year and julian day
  fire_days_short <- subset(fire_days_count, fire_days_count$YEAR < 2014) ##restricting fire records to climate file years
  #FWI_short <- subset(FWI_dat, FWI_dat$Year > 1991) #restricting climate data to fire history records
  merge_col <- FWI_dat$ID
  FWI_fire_merge <- join(FWI_dat, fire_days_short, type="left") ##Merging based on unique date id
  FWI_fire_number <- FWI_fire_merge[,c(1,3,7)] #pulling out FWI and number of fires
  FWI_fire_number[is.na(FWI_fire_number)] <- 0 #converting NAs to 0, meaning 0 fires
  plot(FWI_fire_number[,2], FWI_fire_number[,3], main =ign_types[i], xlab = "dailyFWI", ylab = "noFires") #plotting FWI against no of fires just to look at pattern
  igns_list[[i]] <- FWI_fire_number 
}
```

Summary

```{r Possion GLM}
summary(igns_list[[1]])
Lightning<-as.data.frame(igns_list[[1]])
POIS_mod_L <- glm(No_FIRES~FWI,data=Lightning, family="poisson")
summary(POIS_mod_L)

summary(igns_list[[2]])
Human<-as.data.frame(igns_list[[2]])
POIS_mod_H <- glm(No_FIRES~FWI,data=Human, family="poisson")
summary(POIS_mod_H)

```

## Plotting social climate fire logs against empirical

```{r social climate fire}
Simulated_Fire <- read.csv("F:/Rancher_Sims_Full_Landscape/interior_hist_ncar_3.18.2025/socialclimatefire-summary-log.csv")
landis_fire_days<- Simulated_Fire %>%
  dplyr::rename(YEAR = SimulationYear)%>%
  mutate(YEAR=YEAR+1999)%>%
  group_by(YEAR)%>%
  select(YEAR, TotalBurnedSitesLightning)

sorted_counts <- sort(landis_fire_days$TotalBurnedSitesLightning, decreasing = FALSE)
color_map <- setNames(reds, sorted_counts)
bar_colors <- color_map[as.character(landis_fire_days$TotalBurnedSitesLightning)]

colnames(landis_fire_days)<-c("YEAR", "No_Fires")
barplot(landis_fire_days$No_Fires, main ="No. of fires/yr LANDIS Simulated",col=bar_colors,names.arg=landis_fire_days$YEAR)
barplot(l_fire_days_sort_count$COUNT, main ="No. of lightning ign/yr Empirical",col=bar_colors,names.arg=l_fire_days_sort_count$YEAR)

landis_fire_days_filtered <- landis_fire_days[landis_fire_days$YEAR >= 2000 & landis_fire_days$YEAR <= 2014, ]
l_fire_days_sort_count_filtered <- l_fire_days_sort_count[l_fire_days_sort_count$YEAR >= 2000 & l_fire_days_sort_count$YEAR <= 2014, ]

# Merge the two datasets by the year
combined_data_noFires <- merge(landis_fire_days_filtered, l_fire_days_sort_count_filtered, 
                       by="YEAR",
                       suffixes = c("_LANDIS", "_LIGHTNING"))

#woah
barplot(
  height = rbind(combined_data_noFires$No_Fires, combined_data_noFires$COUNT), 
  beside = TRUE,  
  col = c("red", "blue"),  
  names.arg = combined_data_noFires$YEAR,  
  main = "Number of fires",
  legend.text = c("LANDIS Simulated", "Lightning Empirical"),  
  args.legend = list(x = "topright")
)
```

## Simulated AAB against empirical 

```{r}
l_fire_aab <- l_fire_days_sort %>%
  group_by(FIRE_YEAR) %>%
  dplyr::summarise(COUNT = n(), AAB = sum(FIRE_SIZE/2.471, na.rm= TRUE))
colnames(l_fire_aab) <- c("YEAR", "COUNT", "AAB")
barplot(l_fire_aab$AAB, main ="Annual Area Burned Observed (Ha)",col=reds,names.arg=l_fire_aab$YEAR)

#landis
landis_aab<- Simulated_Fire %>%
  dplyr::rename(YEAR = SimulationYear)%>%
  mutate(YEAR=YEAR+1999)%>%
  group_by(YEAR)%>%
  mutate(AAB = TotalBurnedSitesLightning*4)%>%
  select(YEAR, AAB)
barplot(landis_aab$AAB, main ="Annual Area Burned Simulated (Ha)",col=reds,names.arg=landis_aab$YEAR)


## filter and join
landis_aab_filtered <- landis_aab[landis_aab$YEAR >= 2000 & landis_aab$YEAR <= 2014, ]
l_fire_aab_filtered <- l_fire_aab[l_fire_aab$YEAR >= 2000 & l_fire_aab$YEAR <= 2014, ]

# Merge the two datasets by the year
combined_data_aab <- merge(landis_aab_filtered, l_fire_aab_filtered, 
                       by = "YEAR", 
                       suffixes = c("_LANDIS", "_LIGHTNING"))
#shoot
barplot(
  height = rbind(combined_data_aab$AAB_LANDIS, combined_data_aab$AAB_LIGHTNING), 
  beside = TRUE,  
  col = c("red", "blue"),  
  names.arg = combined_data_aab$YEAR,  
  main = "Annual Area Burned (Ha)",
  legend.text = c("LANDIS Simulated", "Lightning Empirical"),  
  args.legend = list(x = "topright")
)
```

# Probability of establishment with a new climate library (Historic sim)

```{r}
library(viridis)
library(terra)
Pest_graphing_matrix<-read.csv("F:/Rancher_Sims_Full_Landscape/interior_hist_ncar_3.18.2025/DGS-prob-establish-log.csv")%>%mutate(Year=Time+2014)
Pest_graphing_matrix_someSPP <- Pest_graphing_matrix %>%
  filter(!SpeciesName %in% c("TotalBiomass", "Turfmoss", "Sphagnum", 
                         "Feathermoss", "ShrubBirch", "Willow", "Alder", "Tamarack", "BalsamPoplar"))


climRegions_colors <- brewer.pal(10, "Paired")
est_plt <- Pest_graphing_matrix_someSPP %>%
  mutate(EcoRegion=as.factor(ClimateRegion))%>%
  group_by(ClimateRegion)%>%
ggplot(aes(x = Year, y = AvgProbEst, color = EcoRegion)) +
  #geom_smooth(se=FALSE)+
  geom_line(linewidth = 1, alpha = 1) +
  ylab("Probability of Establishment") +
  scale_x_continuous(limits = c(2015, 2100), breaks = seq(2015, 2100, 25), expand = c(0, 0)) +
  scale_color_manual(
    values = climRegions_colors,
    labels = function(x) paste0("eco ", x)
  ) +
  facet_wrap(~SpeciesName) +
  guides(color = guide_legend(override.aes = list(shape = 16))) +
  theme_bw() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14),
    strip.text = element_text(size=8),
    strip.placement = "top",
    axis.title.y = element_text(size = 20, margin = margin(r = 20)),
    axis.title.x = element_text(size = 18, margin = margin(r = 20), hjust=.47, vjust=-0.05),
    panel.spacing.y = unit(2, "lines"),
    panel.grid.major = element_line(color = "gray80", size = 0.5),
    panel.grid.minor = element_line(color = "gray90", size = 0.3),
    axis.text.x = element_text(color = "black"),
    axis.text.y = element_text(color = "black")
  )
est_plt

bounds <- vect("F:/Rancher_Sims_Full_Landscape/input/FullLandscapeV3_082722.shp")
ecoRaster <- rast("F:/Rancher_Sims_Full_Landscape/input/AK_ClimateMap_10_Regions.tif")
ecoRaster <- project(ecoRaster, "EPSG:3338")
ecoRaster <- mask(ecoRaster, bounds)
plot(ecoRaster, col = c("black",climRegions_colors))
```

# Plotting community biomass against FIA


